Modelling low Mach number stellar hydrodynamics with MAESTROeX

### 2019-08-29 astronum_2019_MAESTROeX/astronum-maestroex.tex
add in lots of citations

Sound-proof methods for modelling low Mach number flows can take a variety of forms. One technique is to use preconditioners in order to reduce the stiffness of the equations, allowing larger timesteps to be used CITE. Another technique is to modify the flow parameters, artificially boosting the speed of the flow without changing its behavior so that the system evolves faster and fewer timesteps are required CITE.

The technique that we use in \maestroex~is to modify the fluid equations themselves so as to filter out the soundwaves. This is a similar approach to the incompressible CITE and anelastic CITE approximations, however the approximation that we use (variously called the low Mach number approximation  CITE and the generalized pseudo-incompressible approximation CITE) allows background stratification and large pressure perturbations due to heating and changes in composition. This is achieved by decomposing the pressure into a 1d hydrostatic base state, $p_0 = p_0(r, t)$, and a dynamic pressure perturbation, $\pi = \pi(r, \bm{x}, t)$, such that $p(r, \bm{x}, t) = p_0 + \pi$. In order for the approximation to hold, $|\pi|/p_0 = O(\mathrm{Ma}^2)$. The low Mach number fluid equations are given by 
\begin{align}
    \pd{\left(\rho X_k\right)}{t} &= - \nabla\cdot\left(\rho X_k \Ub \right) + \rho\dot{\omega}_k, \\
    \pd{\Ub}{t} &= - \Ub\cdot\nabla\Ub - \frac{\beta_0}{\rho}\nabla\left(\frac{\pi}{\beta_0}\right) - \frac{\rho - \rho_0}{\rho}\bm{e}_r,\\
    \pd{\left(\rho h\right)}{t} &= -\nabla\cdot\left(\rho h \Ub\right) + \md{p_0}{t} + \rho H_{\mathrm{nuc}},
\end{align}
where $\rho$, $\Ub$ and $h$ are the mass density, fluid velocity and specific enthalpy, $X_k$ and $\dot{\omega}_k$ are the species mass fraction and production rate of species $k$, and $H_{\mathrm{nuc}}$ is the energy release per time per unit mass. From the base state pressure, we can define a 1d base state density, $\rho_0 = \rho_0(r, t)$, which represents the lateral average and is in hydrostatic equilibrium with $p_0$:
\begin{equation}
    \nabla p_0 = -\rho_0 g \bm{e}_r,
\end{equation}
where $g = g(r,t)$ is the magnitude of the gravitational acceleration, and $\bm{e}_r$ is the radial unit vector. The background stratification is captured by introducing a buoyancy-like term, $\beta_0$. It is defined as 
\begin{equation}
    \beta_0(r,t) = \rho_0(0,t) \exp\left(\int_0^r dr'\, \frac{1}{\bar{\Gamma}_1p_0}\pd{p_0}{r'} \right),
\end{equation}
where $\bar{\Gamma}_1$ is the lateral average of the first adiabatic exponent, $\Gamma_1 = d(\ln p)/d(\ln \rho)|_s$, and $s$ is the entropy. The equations are closed by casting the equation of state as a velocity divergence constraint. This is done by taking the Lagrangian derivative of the EoS for pressure as a function of the thermodynamic variables, substituting in the equations of motion for mass and energy, and requiring that the pressure is described by a function of $r$ and $t$ based on the condition of hydrostatic equilibrium. Details of this derivation can be found in CITE. The constraint is given by 
\begin{equation}
    \nabla\cdot\left(\beta_0\Ub\right) = \beta_0 \left(S - \frac{1}{\bar{\Gamma}_1 p_0}\pd{p_0}{t} \right).
\end{equation}
Here, $S$ is an expansion term which describes local compressibility effects due to changes in composition and heating from reactions:
\begin{equation}
    S = -\sigma \sum_k \xi_k\dot{\omega}_k + \frac{1}{\rho p_\rho}\sum_k p_{X_k}\dot{\omega}_k + \sigma H_{\mathrm{nuc}},
\end{equation}
where we define 
\begin{align*}
    p_{X_k} \equiv \left.\pd{p}{X_k}\right|_{\rho,T,X_{j,j\neq k}},\qquad
     \xi_k&\equiv \left.\pd{h}{X_k}\right|_{p, T,X_{j,j\neq k}},\qquad
     p_\rho\equiv \left.\pd{p}{\rho}\right|_{T, X_k},\\
     \sigma \equiv \frac{p_T}{\rho c_p p_\rho}, \qquad
     p_T&\equiv \left.\pd{p}{T}\right|_{\rho, X_k} \quad\mathrm{and}\quad 
     c_p\equiv \left.\pd{h}{T}\right|_{p, X_k}.
\end{align*}

@@

Prior to \maestroex, we developed the low Mach number code \maestro. Like \maestroex, \maestro~is a block-structured adaptive mesh refinement (AMR) code for modelling low Mach number astrophysical codes. It was developed over a series of papers CITE. The system of low Mach number equations are solved using an explicit Godunov approach for the advection, a stiff ODE solver for the reactions (VODE, CITE), and multigrid linear solvers for the pressure projection steps. Strang spitting CITE is used to integrate the thermodynamic variables, a second order projection method to integrate the velocity subject to the divergence constraint, and a velocity splitting scheme to hydrodynamically evolve the base state. The original \maestro~code was developed using the pure Fortran 90 \fboxlib~software framework CITE; \maestroex~instead uses the C++/Fortran 90 \amrex~framework CITE.

\maestro~has been used to model a number of astrophysical systems, including convection in white dwarfs prior to type Ia supernovae CITE, convection in massive stars CITE and type I X-ray bursts CITE.

@@

\maestroex~is part of the \amrexastro~suite of open source adaptive mesh refinement hydrodynamics codes for astrophysical flows. Other codes in this family include \castro, an astrophysical radiation hydrodynamics simulation code, and \nyx, an N-body hydrodynamics cosmological simulation code. All three codes are developed using the \amrex~software framework, and share much in common in terms of their development, structure and numerical methods. Developers of the different codes work closely together (in fact many of the developers work on more than one of the codes), with their shared structure and framework meaning that new features developed in one code can easily be replicated in the others. Examples of this include the SDC solvers CITE and GPU port that have recently been implemented in \castro, and are currently in the process of being implemented in \maestroex.

>>>>

Sound-proof methods for modelling low Mach number flows can take a variety of forms. One technique is to use preconditioners in order to reduce the stiffness of the equations, allowing larger timesteps to be used \cite{Miczek2014,Barsukow2016}. Another technique is to modify the flow parameters, artificially boosting the speed of the flow without changing its behavior so that the system evolves faster and fewer timesteps are required \citep{Rempel2005,Hotta2012}. Fully implicit time integration codes, such as those used by \cite{Viallet2011,Viallet2015,Goffrey2016}, can use arbitratily timesteps, so are no longer restricted by the CFL condition.

The technique that we use in \maestroex~is to modify the fluid equations themselves so as to filter out the soundwaves. This is a similar approach to the incompressible \cite{Boussinesq1901} and anelastic \cite{Ogura1962a,Gough1968,Durran1989} approximations, however the approximation that we use (variously called the low Mach number approximation \cite{Day2000,Almgren2006a,Nonaka2010} and the generalized pseudo-incompressible approximation \cite{Vasil2013}) allows background stratification and large pressure perturbations due to heating and changes in composition. This is achieved by decomposing the pressure into a 1d hydrostatic base state, $p_0 = p_0(r, t)$, and a dynamic pressure perturbation, $\pi = \pi(r, \bm{x}, t)$, such that $p(r, \bm{x}, t) = p_0 + \pi$. In order for the approximation to hold, $|\pi|/p_0 = O(\mathrm{Ma}^2)$. The low Mach number fluid equations are given by 
\begin{align}
    \pd{\left(\rho X_k\right)}{t} &= - \nabla\cdot\left(\rho X_k \Ub \right) + \rho\dot{\omega}_k, \\
    \pd{\Ub}{t} &= - \Ub\cdot\nabla\Ub - \frac{\beta_0}{\rho}\nabla\left(\frac{\pi}{\beta_0}\right) - \frac{\rho - \rho_0}{\rho}\bm{e}_r,\\
    \pd{\left(\rho h\right)}{t} &= -\nabla\cdot\left(\rho h \Ub\right) + \md{p_0}{t} + \rho H_{\mathrm{nuc}},
\end{align}
where $\rho$, $\Ub$ and $h$ are the mass density, fluid velocity and specific enthalpy, $X_k$ and $\dot{\omega}_k$ are the species mass fraction and production rate of species $k$, and $H_{\mathrm{nuc}}$ is the energy release per time per unit mass. From the base state pressure, we can define a 1d base state density, $\rho_0 = \rho_0(r, t)$, which represents the lateral average and is in hydrostatic equilibrium with $p_0$:
\begin{equation}
    \nabla p_0 = -\rho_0 g \bm{e}_r,
\end{equation}
where $g = g(r,t)$ is the magnitude of the gravitational acceleration, and $\bm{e}_r$ is the radial unit vector. The background stratification is captured by introducing a buoyancy-like term, $\beta_0$. It is defined as 
\begin{equation}
    \beta_0(r,t) = \rho_0(0,t) \exp\left(\int_0^r dr'\, \frac{1}{\overline{\Gamma}_1p_0}\pd{p_0}{r'} \right),
\end{equation}
where $\overline{\Gamma}_1$ is the lateral average of the first adiabatic exponent, $\Gamma_1 \equiv d(\ln p)/d(\ln \rho)|_s$, and $s$ is the entropy. The equations are closed by casting the equation of state as a velocity divergence constraint. This is done by taking the Lagrangian derivative of the EoS for pressure as a function of the thermodynamic variables, substituting in the equations of motion for mass and energy, and requiring that the pressure is described by a function of $r$ and $t$ based on the condition of hydrostatic equilibrium. Details of this derivation can be found in \cite{Almgren2006a,Almgren2006b}. The constraint is given by 
\begin{equation}
    \nabla\cdot\left(\beta_0\Ub\right) = \beta_0 \left(S - \frac{1}{\overline{\Gamma}_1 p_0}\pd{p_0}{t} \right).
\end{equation}
Here, $S$ is an expansion term which describes local compressibility effects due to changes in composition and heating from reactions:
\begin{equation}
    S = -\sigma \sum_k \xi_k\dot{\omega}_k + \frac{1}{\rho p_\rho}\sum_k p_{X_k}\dot{\omega}_k + \sigma H_{\mathrm{nuc}},
\end{equation}
where we define 
\begin{align*}
    p_{X_k} \equiv \left.\pd{p}{X_k}\right|_{\rho,T,X_{j,j\neq k}},\qquad
     \xi_k&\equiv \left.\pd{h}{X_k}\right|_{p, T,X_{j,j\neq k}},\qquad
     p_\rho\equiv \left.\pd{p}{\rho}\right|_{T, X_k},\\
     \sigma \equiv \frac{p_T}{\rho c_p p_\rho}, \qquad
     p_T&\equiv \left.\pd{p}{T}\right|_{\rho, X_k} \quad\mathrm{and}\quad 
     c_p\equiv \left.\pd{h}{T}\right|_{p, X_k}.
\end{align*}

@@

Prior to \maestroex, we developed the low Mach number code \maestro. Like \maestroex, \maestro~is a block-structured adaptive mesh refinement (AMR) code for modelling low Mach number astrophysical codes. It was developed over a series of papers \cite{Almgren2006a,Almgren2006b,Almgren2008a,Zingale2009,Nonaka2010}. The system of low Mach number equations are solved using an explicit Godunov approach for the advection, a stiff ODE solver for the reactions (VODE, CITE), and multigrid linear solvers for the pressure projection steps. Strang spitting CITE is used to integrate the thermodynamic variables, a second order projection method to integrate the velocity subject to the divergence constraint, and a velocity splitting scheme to hydrodynamically evolve the base state. The original \maestro~code was developed using the pure Fortran 90 \fboxlib~software framework \cite{Zhang2016}; \maestroex~instead uses the C++/Fortran 90 \amrex~framework \cite{Zhang2019}.

\maestro~has been used to model a number of astrophysical systems, including convection in white dwarfs prior to type Ia supernovae \cite{Zingale2011,Nonaka2011,Malone2014a,Zingale2013,Jacobs2016}, convection in massive stars \cite{Gilet2013} and type I X-ray bursts \cite{Malone2011,Malone2014,Zingale2015}.

@@

\maestroex~is part of the \amrexastro~suite of open source adaptive mesh refinement hydrodynamics codes for astrophysical flows. Other codes in this family include \castro, an astrophysical radiation hydrodynamics simulation code \cite{Almgren2010}, and \nyx, an N-body hydrodynamics cosmological simulation code \cite{Almgren2013}. All three codes are developed using the \amrex~software framework, and share much in common in terms of their development, structure and numerical methods. Developers of the different codes work closely together (in fact many of the developers work on more than one of the codes), with their shared structure and framework meaning that new features developed in one code can easily be replicated in the others. Examples of this include the SDC solvers \cite{Zingale2019} and GPU port that have recently been implemented in \castro, and are currently in the process of being implemented in \maestroex.
